---
title: "A quickstart guide to GEMINI"
author: "Mahdi Zamanighomi and Sidharth Jain"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{QuickStart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This is a tl;dr of the `GEMINI` manual, but should allow an experienced user to quickly begin to use the many features of `GEMINI`.

GEMINI follows a basic workflow:

- Create Input (`gemini_create_input`)
- Initialize Model (`gemini_initialize`)
- Perform Inference (`gemini_inference`)
- Score Interactions (`gemini_score`)

```{r}
library("gemini")
data("example-bigPapi", package = "gemini")
```

### Input
GEMINI takes a counts matrix as follows:
```{r, echo = F}
knitr::kable(head(counts[,1:5]), caption = "Counts matrix", align = 'l')
```

GEMINI also requires sample/replicate annotation and guide/gene annotation.
```{r, echo = F}
knitr::kable(head(sample.replicate.annotation), caption = "Sample/replicate annotations")
knitr::kable(head(guide.annotation), caption = "Guide/gene annotation")
```

These can be used to create a `gemini.input` object using the `gemini_create_input` function.
```{r}
Input <- gemini_create_input(counts.matrix = counts,
                    sample.replicate.annotation = sample.replicate.annotation,
                    guide.annotation = guide.annotation,
                    ETP.column = 'pDNA', # this can also be specified by column index
                    gene.column.names = c("U6.gene", "H1.gene"),
                    sample.column.name = "samplename",
                    verbose = T)
```

### Pre-processing
GEMINI requires log-fold changes as an input, which are calculated using the `gemini_calculate_lfc` function.
```{r}
Input %<>% gemini_calculate_lfc(normalize = T, 
                                CONSTANT = 32)
```

### Initialization and Inference
To initialize the model, a `gemini.model` object is created using the `gemini_initialize` function.

To note, at least one negative control gene **MUST** be specified here, and that the guides targeting this negative control gene **MUST** be paired with all other genes. We use CD81 in this example.

Also to note, the `pattern_split` argument must describe a separator used in the rownames of the counts.matrix. For example, a semi-colon (;) is used in the Big Papi data and therefore specified here.
```{r}
Model <- gemini_initialize(Input = Input, 
                  nc_gene = "HPRT intron", 
                  pattern_join = '-',
                  pattern_split = ';', 
                  cores = 1,
                  verbose = T)
```

Inference is performed with the `gemini_inference` function.

```{r}
Model %<>% gemini_inference(cores = 1,
                            verbose = F)
```

Convergence is visualized with the `gemini_plot_mae` function.

```{r}
gemini_plot_mae(Model)
```

### Scoring and Visualization
To score genetic interactions, use the `gemini_score` function.

To note, at least one positive control gene (`pc_gene`) should be specified to remove interactions involving individually lethal genes. If no positive control is explicitly specified, lethality is estimated as described in Zamanighomi *et al* ?.

Additionally, non-interacting gene pairs (`nc_pairs`) should be specified for the calculation of p-values and false discovery rates (FDRs). If not specified, only scores are calculated.

```{r}
# Use other negative controls or known non-interacting gene pairs as nc_pairs
nc_pairs <- grep("6T|CD81", rownames(Model$s), value = T)
# Some nc_pairs...
head(nc_pairs, n = 5)

Score <- gemini_score(Model = Model,
             pc_gene = "EEF2",
             nc_pairs = nc_pairs)
```

Significant interactions can be identified through the FDR and p-value slots in the `Score` object. For example, the strongest interactions in the "A549" cell line can be found here:

```{r, echo = F}
knitr::kable(Score$fdr_strong[order(Score$fdr_strong[,"A549"], decreasing = F)[1:10],], caption = "FDRs for top 10 interactions for A549")
```

To visualize these interactions, we can use the `gemini_boxplot` function. For example, in BRCA2-PARP1:

```{r, eval = F}
gemini_boxplot(Model = Model, 
               gene.column.names = c("U6.gene", "H1.gene"), 
               g = "BRCA2",
               h = "BRCA1",
               nc_gene = "HPRT intron",
               sample = "A549",
			   show_inference = T, 
			   identify_guides = T
			   )
```


We can see that GEMINI makes adjustments to the individual gene effects of both BRCA2 and PARP1, and adjusts the sample-independent values of each to account for variation in the screen. Guides with the least adjustment to x (grey) are considered to have the least variation within the screen.

### Summary

GEMINI can be run on any counts matrix from a pairwise screen. GEMINI computes log fold changes and infers sample-dependent and sample-independent effects. Then, GEMINI calculates interaction strength and significance. 

See Zamanighomi et al 

